\chapter{Acknowledgements}\label{acknowledgments}

In spring 2017, the two of us met up as friends. We discussed all sorts of things, one of them being the history of English. Both of us voiced some frustration over the suitability of the materials available for courses on this area of English linguistics. We felt that although there are a lot of materials, they do not necessarily fit the format of our individual courses, so much so that Míša professed that ``Sometimes I wonder if it might be easier just to write my own book''. After some silence, we looked at each other, and it was clear to us what would happen. And here we are, five years later, with a baby of a textbook.

Many individuals have helped us. First of all, we would like to thank the students we have taught for the History of the English Language at Aarhus University between 2017--2021, some of whom served as guinea pigs for one chapter while others for the entire book. The same goes for students of the History of the English Language and of Old English at Konstanz from 2017 to 2021. To us, our students were one of the best reviewing committees we could have asked for. Among these studens we would particularly like to thank Dominik Dolt, Jonathan Förderer, Robert Langen, Philipp Weiss, and Laura Weisshaar.

Many academics have read various drafts of this book, either in its entirety or individual chapters. First and foremost, we would like to thank the editors of the Language Science Press series Textbooks in Language Sciences, Stefan Müller and Martin Haspelmath, and Martin in particular, who provided us with very useful feedback while giving us the freedom we needed to make this book happen. Sebastian Nordhoff has our thanks and gratitude for all his work on the mountain of the final typesetting work. We were also fortunate to have an army of eleven reviewers, all of whom contributed to the process of making this book take its current shape. Melanie Röthlisberger has our special thanks. Other friendly souls who were kind enough to spend their time on the book and/or discuss it with us in various ways include Aaron Ecay, Gry Faurholt, James Harland, Frederik Hartmann, Anna Jespersen, Helle Kaalund Tornbo, Mads Kristian Andersen, Anne Mette Nyvad, Mirjam Schmalz, Nicole Tamer, Krestina Vendelbo Christensen, Fiona Thomas, Sten Vikner, Michael Walkden, and Daniel Watson.

This book is dedicated to our students: past, present, and future. The future history of the English language is in your hands!